---
title: ""
author: "Bayansulu Tulepbayeva and Alua Birgebayeva"
date: "Case Study 1: March 5, 2025"
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
---

_Your written short report should be the first thing your reader encounters in your case study document. While you may choose to label each of the required components/sections of the short report with its own header (see below for an example of creating a header in Quarto), you do not need to do so provided that all of the required information is included. Your report should be self-contained and written in such a way that a quantitatively inclined friend (who has taken or is taking SDS 291) could follow what you did without necessarily knowing anything about the U.S. presidential election of 2000._ 

_As you write your report, you may wish to reference the guide to typesetting regression lines in Quarto using LaTeX (linked at the top of our class Moodle page), the Quarto help page for formatting documents using Markdown (https://quarto.org/docs/authoring/markdown-basics.html), and the Quarto help page for customizing the output from executed code chunks (https://quarto.org/docs/computations/execution-options.html)._

### Introduction


### Data & Methodology

We analyzed 67 Florida counties using data from the Sleuth2 package. Since Palm Beach County is suspected to be an outlier, we excluded it from our regression model to avoid bias.

To estimate how many votes Buchanan should have received in Palm Beach, we built a linear regression model predicting Buchanan votes based on Bush votes in the other counties. The equation of our model is:
```{r}
#| include: false

# Loading necessary packages
library(tidyverse)
library(Sleuth2)      # the package containing the data for the case study
library(kableExtra)   # for creating nicely formatted tables in Quarto

# Loading the case study data
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")

```

### Regression Model 

```{r}
#| echo: false
#| message: false
#| warning: false

# Fit the linear model
model <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)

# Display regression summary
summary(model)

```

```{r}
# Format and display the regression summary in a clean table
library(kableExtra)

# Extract regression coefficients
regression_table <- summary(model)$coefficients |> 
  kbl(col.names = c("Estimate", "Std. Error", "t value", "Pr(>|t|)"), 
      align = "c", booktabs = TRUE) |> 
  kable_classic(full_width = FALSE, latex_options = c("HOLD_position"))

# Show the table
regression_table

```

### Prediction for Palm Beach County 

```{r}
# Get Bush votes in Palm Beach County
pb_bush_votes <- election |> filter(County == "Palm Beach") |> pull(Bush2000)

# Predict Buchanan votes for Palm Beach (95% prediction interval)
predicted_pb_buchanan <- predict(model, newdata = data.frame(Bush2000 = pb_bush_votes), interval = "prediction", level = 0.95)

# Show the prediction interval
predicted_pb_buchanan

predicted_pb_df <- data.frame(
  County = "Palm Beach",
  Predicted_Buchanan_Votes = predicted_pb_buchanan[1],
  Lower_Bound = predicted_pb_buchanan[2],
  Upper_Bound = predicted_pb_buchanan[3]
)

predicted_pb_df |> 
  kbl(col.names = c("County", "Predicted Buchanan Votes", "Lower Bound (95% CI)", "Upper Bound (95% CI)"), 
      align = "c", booktabs = TRUE) |> 
  kable_classic(full_width = FALSE, latex_options = c("HOLD_position"))

```

### Actual vs. Predicted Votes & Miscast Estimate

```{r}
# Get actual Buchanan votes in Palm Beach
actual_pb_buchanan <- election |> filter(County == "Palm Beach") |> pull(Buchanan2000)

# Print actual Buchanan votes
actual_pb_buchanan

```

### Estimating the Midcast votes 

```{r}
# Estimate miscast votes (Actual Buchanan votes - Upper Bound of Prediction Interval)
miscast_votes <- actual_pb_buchanan - predicted_pb_buchanan[3]

# Print estimated miscast votes
miscast_votes

```
 
 
### Results 

```{r}
# Create a summary table for the report
miscast_summary_df <- data.frame(
  County = "Palm Beach",
  Actual_Buchanan_Votes = actual_pb_buchanan,
  Predicted_Upper_Bound = predicted_pb_buchanan[3],
  Estimated_Miscast_Votes = miscast_votes
)

# Display as a well-formatted table
library(kableExtra)
miscast_summary_df |> 
  kbl(col.names = c("County", "Actual Buchanan Votes", "Upper Bound Prediction", "Estimated Miscast Votes"), 
      align = "c", booktabs = TRUE) |> 
  kable_classic(full_width = FALSE, latex_options = c("HOLD_position"))

```

_When reporting the results of your regression analysis, I would like you to write out a mathematical description of the main/final population model for the mean. Here is an example of what that might look like:_

Let $mort_i$ denote the heart disease mortality rate in country $i$ (measured in deaths per 1,000 individuals) and $wine_i$ denote the consumption of wine (measured in liters per person per year). Our final linear regression model is $$E[mort_i | wine_i] = \beta_0 + \beta_1\left(wine_i\right).$$

_You do not need to write out the fitted regression line, but you should (at a minimum) provide a table summarizing the estimated coefficients of the model and their corresponding standard errors, like the one shown below. A print-out from the `summary()` command alone is not sufficient._

```{r}
#| echo: false


```

# R Appendix

_Copy and paste all code that you used for your case study into one chunk at the end of your written report. Before submitting your case study, take one final look at the R Appendix and make sure that all code is clearly visible. If you see a line running off the side of the PDF, please split the code over multiple lines using a linebreak. The header at the top of this template should ensure that lines of code are automatically wrapped in your final document._

```{r}
#| message: FALSE
#| warning: FALSE


```
